The LSST-DESC 3x2pt Tomography Optimization Challenge

### 2021-08-17 paper/paper.tex
update malz contrib
changes from Danielle

The \emph{3x2pt} method has become a standard tool for performing this combination.
In this method, two-point correlations are computed among and between two samples, the shapes of 
background (source) galaxies and the locations of foreground galaxies, which trace foreground
dark matter haloes.  The three combinations (source-source, source-lens, and lens-lens) are
measured in either Fourier or configuration space, and can be predicted from a combination of 
perturbation theory and simulation results.  The method has been used in the Dark Energy Survey, DES, 
\citep{des-3x2pt}, and to combine the Kilo-Degree 
Survey, KiDS with spectroscopic surveys \citep{kids-3x2pt, kids_gama, kids_2df}.

@@

We are free to assign galaxies to different tomographic bins in any way we wish; changing the choice
can potentially affect the ease with which we can calibrate
the bins, but any choice can lead to correct results. The bins need not even be purely tomographic: we can happily correlate bins with number density peaks at different redshifts if that is useful.
We should choose, then, an assignment algorithm that maximises the constraining power for a science
case of interest.  This challenge explores such algorithms.

@@

This paper is part of preparations for the analysis of Legacy Survey of Space and Time (LSST) data at the Rubin Observatory, being run by the Dark Energy Science Collaboration (DESC).
In it, we discuss and evaluate the related challenge
of using a limited set of colour bands to make those assignments, motivated by requirements of using the
metacalibration method for galaxy shape measurement to limit biases.  
A similar question, with different motivations, was explored in \citet{jain}, who found that 3-band 
$gri$ tomography could be effective for tomographic selection, a result we will echo here in a new context: in this paper we describe the results of a 
challenge using simulated Rubin-like data with a range of methods submitted by different teams.  We explore how many tomographic bins we can effectively generate using only a subset of bins, and compare methods submitted to the challenge.

@@

In the challenge, participants assigned simulated galaxies to tomographic bins,
using only their (g)riz bands and associated errors, and the galay size.  Their goal was to maximize the cosmological
information in the final split data set; this is generally achieved by separating different bins by 
redshift as much as possible.  Particpants were free to optimize or simply select nominal (target) 
redshift bin edges from their training sample, but in either case had to assign galaxies into the 
different bins.

@@

The challenge was open to any entrants, not just those already involved in DESC, but it was
advertised through Rubin, and so is perhaps best described as semi-public.
Participants developed and tested their methods locally but submitted code to be run as part of a central
combined analysis at the National Energy Research
Supercomputing Center (NERSC), where the final tests and metric evaluation were conducted.  No prize was offered apart from recognition.

@@

We ran challenge entries on two data sets, CosmoDC2 and Buzzard, each of which is split into training, 
validation, and testing components.  In each case participants were supplied with the training and 
validation data, and the testing data was kept secret for final evaluation. In each case the three
data sets were random sub-selections of the full data sets, with the training and testing sets 25\% of 
the full data each and the validation 50\%. In practice many entrants trained on a subset of
this, so to make the final evaluation phase tractable we trained on 1M objects for both data sets.

@@

Metric 2 uses a figure of merit (FOM) based on that presented by the Dark Energy Task Force (DETF) \citep{detf}.  We 
use the inverse area of a Fisher Matrix ellipse, which approximates the size in two axes 
of the posterior PDF one would obtain on analysing the spectra:
\begin{align}
    F &= \left( \frac{\partial \mu}{\partial \theta} \right)^T C^{-1} \left( \frac{\partial \mu}{\partial \theta} \right), \\
    S &= \frac{1}{2 \pi \sqrt{\det{([F^{-1}]_{p_1, p_2})}}}
\label{eq:fom}
\end{align}
where $[F^{-1}]_{p_1, p_2}$ extracts the 2x2 matrix for two parameters.  We use both the original
DETF variant in which $p_1 = w_0$ and $p_2 = w_a$, representing the constraining power on dark energy 
evolution, and a version in which $p_1 = \Omega_c$ and $p_2 = \sigma_8$, representing constraining
power on overall cosmic structure amplitude and its evolution. In each case the complete parameter space consists
of the eight $w_0w_aCDM$ cosmological parameters. This simplified measure does not include any nuisance
parameters modelling systematic errors in the analysis (most importantly the galaxy bias parameters in the clustering 
spectra), and like all Fisher analyses approximates posterior constraints as Gaussian.

@@

\subsection{Results Overview}
The DETF results for each method are shown in Table \ref{tab:cosmodc2}.  Complete results for all metrics
for the nine-bin scenario only are shown in appendix tables \ref{tab:full_dc2} and \autoref{tab:full_buzz}.

@@

\begin{figure}[htbp]
	% \centering
	\includegraphics[width=1\linewidth]{results/nzbinned_dc2_funbins.pdf}
	\caption{The tomographic $n(z)$ values generated by the winning method in the challenge, {\sc funbins},
		generating nine bins with the CosmoDC2 data.
		The upper panel shows results for the method applied to $riz$ bands, and the middle on $griz$. The
		increased spread and overlap of the bins in the middle panel decreases the overall constraining
		power and signal-to-noise.  The lower panel is for comparison and is not part of the challenge,
		and illutrates results using all the $ugrizy$ bands.}
	\label{fig:funbin_nz}
\end{figure}

@@

These results suggest that, at least in our idealised scenario of perfect training sets and given the limitations of our simulations, restricted photometry does does not necessarily catastrophically limit our ability to use tomography.
In both the CosmoDC2 and Buzzard data sets,
even the $riz$ bands alone are sufficient to split objects into nine reasonably separated
tomographic bins.  Once bins become this narrow the primary analysis concern will become the calibration 
of overall bin $n(z)$ values, rather than initial tomography, so we consider this to be reassuring.

@@

Several of the methods with optimized target bin edges break this
barrier and achieve very high scores.  This included ZotNet, ZotBin,  and JaxResNet.  The
Stacked Generalization method similarly included manually tuned bin edges to maximize
score for this scenario (and indeed its ensemble combination method could be applied to
combine the other successful methods). The typical 15--20\% improvement in FoM scores when optimizing
nominal bin edges makes clear that this easy win should be compelling.

@@

\subsection{Author Contributions}
Zuntz led the project, ran the challenge, wrote most paper text, and made plots.
Lanusse submitted methods, made plots, and contributed to metrics, infrastructure, analysis, and conclusions.
Malz and Wright submitted methods, made plots, and contributed to analysis and conclusions.
Slosar submitted methods, wrote initial infrastructure, and assisted with challenge design.
Authors Abolfathi to Teixeira submitted methods to the challenge.
Subsequent authors generated the data used in the challenge and contributed to infrastructure used in the project.

>>>>

The \emph{3$\times$2pt} method has become a standard tool for performing this combination.
In this method, two-point correlations are computed among and between two samples, the shapes of 
background (source) galaxies and the locations of foreground (lens) galaxies, which trace foreground
dark matter haloes.  The three combinations (source-source, source-lens, and lens-lens) are
measured in either Fourier or configuration space, and can be predicted from a combination of 
perturbation theory and simulation results.  The method has been used in the Dark Energy Survey, DES, 
\citep{des-3x2pt}, and to combine the Kilo-Degree 
Survey, KiDS with spectroscopic surveys \citep{kids-3x2pt, kids_gama, kids_2df}.

@@

We are free to assign galaxies to different tomographic bins in any way we wish; changing the choice
can potentially affect the ease with which we can calibrate
the bins, but any choice can lead to correct cosmology results. The bins need not even be assigned contiguous redshift ranges: we can happily correlate bins with multiple redshifts if that is useful, or by some other galaxy property than redshift entirely.
We should choose, then, an assignment algorithm that maximises the constraining power for a science
case of interest.  This challenge explores such algorithms.

@@

This paper is part of preparations for the analysis to be run by the Dark Energy Science Collaboration (DESC) of Legacy Survey of Space and Time (LSST) data from the Rubin Observatory.
In it, we discuss and evaluate the related challenge
of using a limited set of colour bands to make those assignments, motivated by requirements of using the
metacalibration method for galaxy shape measurement to limit biases.  
A similar question, with different motivations, was explored in \citet{jain}, who found that 3-band 
$gri$ tomography could be effective for tomographic selection, a result we will echo here in a new context: in this paper we describe the results of a 
challenge using simulated Rubin-like data with a range of methods submitted by different teams.  We explore how many tomographic bins we can effectively generate using only a subset of bins, and compare methods submitted to the challenge.

@@

In the challenge, participants assigned simulated galaxies to tomographic bins,
using only their (g)riz bands and associated errors, and the galaxy size.  Their goal was to maximize the cosmological
information in the final split data set; this is generally achieved by separating different bins by 
redshift as much as possible.  Particpants were free to optimize or simply select nominal (target) 
redshift bin edges from their training sample, but in either case had to assign galaxies into the 
different bins.

@@

The challenge was open to any entrants, not just those already involved in DESC, but it was
advertised through Rubin channels, and so is perhaps best described as semi-public.
Participants developed and tested their methods locally but submitted code to be run as part of a central
combined analysis at the National Energy Research
Supercomputing Center (NERSC), where the final tests and metric evaluation were conducted.  No prize was offered apart from recognition.

@@

We ran challenge entries on two data sets, CosmoDC2 and Buzzard, each of which is split into training, 
validation, and testing components.  In each case participants were supplied with the training and 
validation data, and the testing data was kept secret for final evaluation. In each case the three
data sets were random sub-selections of the full data sets, with the training and testing sets 25\% of 
the full data each and the validation 50\%. In practice many entrants trained on a subset of
this, so to make the final evaluation of methods tractable we used 1M objects for both data sets when
computing the metrics presented in this paper.

@@

Metric 2 uses a figure of merit (FOM) based on that presented by the Dark Energy Task Force (DETF) \citep{detf}.  We 
use the inverse area of a Fisher Matrix ellipse, which approximates the size in two axes 
of the posterior PDF one would obtain on analysing the spectra:
\begin{align}
    F &= \left( \frac{\partial \mu}{\partial \theta} \right)^T C^{-1} \left( \frac{\partial \mu}{\partial \theta} \right), \\
    S &= \frac{1}{2 \pi \sqrt{\det{([F^{-1}]_{p_1, p_2})}}}
\label{eq:fom}
\end{align}
where $[F^{-1}]_{p_1, p_2}$ extracts the 2x2 matrix for two parameters.  We use both the original
DETF variant in which $p_1 = w_0$ and $p_2 = w_a$, representing the constraining power on dark energy 
evolution, and a version in which $p_1 = \Omega_c$ and $p_2 = \sigma_8$, representing constraining
power on overall cosmic structure amplitude and its evolution. In each case the complete parameter space consists
of the seven $w_0w_aCDM$ cosmological parameters; we used the parameter combination used natively by the DESC core cosmology library \citep{ccl} ($\Omega_c, \Omega_b, \mathrm{H}_0, \sigma_8, n_s, w_0, w_a$). This simplified measure does not include any nuisance
parameters modelling systematic errors in the analysis (most importantly the galaxy bias parameters in the clustering 
spectra), and like all Fisher analyses approximates posterior constraints as Gaussian.

@@

\subsection{Results Overview}
The DETF results for each method are shown in Table \ref{tab:cosmodc2}.  Complete results for all metrics
for the nine-bin scenario only are shown in appendix Tables \ref{tab:full_dc2} and \autoref{tab:full_buzz}.

@@

\begin{figure}[htbp]
	% \centering
	\includegraphics[width=1\linewidth]{results/nzbinned_dc2_funbins.pdf}
	\caption{The tomographic $n(z)$ values generated by the winning method in the challenge, {\sc funbins},
		generating nine bins with the CosmoDC2 data.
		The upper panel shows results for the method applied to $riz$ bands, and the middle on $griz$.
		The lower panel is for comparison and is not part of the challenge,
		and illustrates results using all the $ugrizy$ bands.
		The increased spread and overlap of the bins as bands are removed decreases the overall constraining
		power and signal-to-noise.
	}
	\label{fig:funbin_nz}
\end{figure}

@@

These results suggest that, at least in our idealised scenario of perfect training sets and given the limitations of our simulations, restricted photometry does not necessarily catastrophically limit our ability to use tomography.
In both the CosmoDC2 and Buzzard data sets,
even the $riz$ bands alone are sufficient to split objects into nine reasonably separated
tomographic bins.  Once bins become this narrow the primary analysis concern will become the calibration 
of overall bin $n(z)$ values, rather than initial tomography, so we consider this to be reassuring.

@@

Several of the methods with optimized target bin edges break this
barrier and achieve very high scores.  This included ZotNet, ZotBin,  and JaxResNet.  The
Stacked Generalization method similarly included manually tuned bin edges to maximize
score for this scenario (and indeed its ensemble combination method could be applied to
combine the other successful methods). The typical 15--20\% improvement in FoM scores when optimizing
nominal bin edges makes clear that re-selecting edges for each science case is worth the time and effort required.

@@

\subsection{Author Contributions}
Zuntz led the project, ran the challenge, wrote most paper text, and made plots.
Lanusse submitted methods, made plots, and contributed to metrics, infrastructure, analysis, and conclusions.
Malz and Wright made plots, and contributed to analysis and conclusions; Wright also submitted methods.
Slosar submitted methods, wrote initial infrastructure, and assisted with challenge design.
Authors Abolfathi to Teixeira submitted methods to the challenge.
Subsequent authors generated the data used in the challenge and contributed to infrastructure used in the project.
